Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-05 sections/eval/api.tex
Update on Overleaf.
fixes

\subsection{Implementing Numpy's Array API in Finch}
In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this and similar APIs for structured data suffer from either incompleteness or inefficiency.
%
These APIs have hundreds of required functions, from mapreduce to slicing and windowing. 
%
Compilers with simpler interfaces (such as TACOs simple einsum) need a massive amount of glue code outside of the normal compilation path to support the full API, leading to a large implementation burden.
%
As the API's were designed for only one structure (Dense), the burden only grows with the introduction of as additional structures such as sparsity, as these structures interact with the complex interface and with each other.
%
%In order to reduce the complexity of interactions between different formats, they're either limited to vectors and matrices or only support tabular representations.
%
Further, implementing each operation on its own is insufficient. Operator fusion is required to avoid asymptotic performance degredation on sparse kernels, as we show in Figure~\ref{fig:fusion}.
%
A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to implement the full array API for structured data with competitive performance.
%
To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
%
This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Physical Interpreter, which executes Finch Logic as a sequence of Finch programs.

@@

\begin{wrapfigure}{R}{0.6\linewidth}
  \scriptsize
\begin{minted}{julia}
  EXPR := table(TENSOR, FIELDS...) |         #TENSOR is an immediate tensor
    ALIAS | FIELD | mapjoin(OP, EXPRS...) |  #OP is a immediate function
    aggregate(OP, INIT, EXPR, FIELDS...) |   #INIT is an immediate init value
    reorder(EXPR, FIELDS...) |               #FIELD is a name for an index
    relabel(EXPR, FIELDS...) |               #ALIAS is a name for a tensor
    reformat(TENSOR, EXPR)                     PLAN := plan(QUERIES..., PRODUCES)
                                              QUERY := query(ALIAS, EXPR)
\end{minted}
\caption{\scriptsize Finch Logic Syntax. }\label{fig:finch_logic_syntax}
\begin{minted}{python}
  sum(A, dims=2) = aggregate(+, relabel(A, i_1, ..., i_d), i_2)
  matmul(A, dims=2) = aggregate(+, mapjoin(*, relabel(A, i, j), relabel(B, j, k)))
  argmin(A) = aggregate(minby, (Inf, len(A)), mapjoin(tuple, relabel(A, i), i), i)
  norm2(A) = aggregate(scaled_plus, (0, 0), mapjoin(tuple, mapjoin(sign, A), mapjoin(abs, A)), fields(A)...)
  where(C, A, B) = mapjoin(ifelse, relabel(C, i), relabel(A, i), relabel(B, i))
  slice(A, i:j) = table(WindowedArray(A, i:j), k)
  cat(A, B) = mapjoin(coalesce, table(PermissiveArray(A), i), table(OffsetArray(B, length(A)), i))
\end{minted}
\caption{\scriptsize Implementations of a few common functions in Finch Logic. Here, \texttt{minby(x, y)} compares \texttt{x[1]} and \texttt{y[1]} and returns the smaller \texttt{x} or \texttt{y}. \texttt{scaled\_plus} rescales whichever argument is smaller, then adds the values with the same scale. \texttt{ifelse(c, a, b)} returns \texttt{a} when \texttt{c} is true,  \texttt{b} otherwise. Finch can understand declared properties and rewrite rules for all of these functions and types, extending these algorithms to operate on structured arguments.} \label{fig:api_extensible}
\end{wrapfigure}
\subsubsection{Finch Logic and the Physical Interpreter}
Finch Logic is used as a common representation to implement and fuse the operations in the Array API.

\paragraph{Finch Logic}
The syntax of Finch Logic is described in Figure~\ref{fig:finch_logic_syntax}.
%
Note that the syntax is designed to be minimal and expressive, taking inspiration from relational algebra with the goal of being able to express most of the operations in the Array API.
%
Our \texttt{mapjoin} represents pointwise application, while \texttt{aggregate} represents reductions.
%
We name the result of each computation with \texttt{query}, and join several expressions in a \texttt{plan}.
%
\texttt{reorder} reorders the indices while \texttt{relabel} relabels them.
%
\texttt{reformat} hints at materialization into its argument.

>>>>

\subsection{Implementing Numpy's Array API in Finch}
In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this and similar APIs for structured data suffer from either incompleteness or inefficiency.
%
These APIs have hundreds of required functions, from mapreduce to slicing and windowing. 
%
Compilers with simpler interfaces (such as TACOs simple einsum) need a massive amount of glue code outside of the normal compilation path to support the full API, leading to a large implementation burden.
%
As the API's were designed for only one structure (Dense), the burden only grows with the introduction of as additional structures such as sparsity, as these structures interact with the complex interface and with each other.
%
%In order to reduce the complexity of interactions between different formats, they're either limited to vectors and matrices or only support tabular representations.
%
Further, implementing each operation on its own is insufficient. Operator fusion is required to avoid asymptotic performance degradation on sparse kernels, as we show in Figure~\ref{fig:fusion}.
%
A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to implement the full array API for structured data with competitive performance.
%
To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
%
This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Physical Interpreter, which executes Finch Logic as a sequence of Finch programs.

@@

\begin{wrapfigure}{R}{0.57\linewidth}
  \scriptsize
  \vspace{-12pt}
\begin{minted}{julia}
  EXPR := table(TENSOR, FIELDS...) |           #TENSOR is a tensor value
    ALIAS | FIELD | mapjoin(OP, EXPRS...) |    #OP is a function
    aggregate(OP, INIT, EXPR, FIELDS...) |     #INIT is an initial value
    reorder(EXPR, FIELDS...) |                 #FIELD is a name for an index
    relabel(EXPR, FIELDS...) |                 #ALIAS is a name for a tensor
    reformat(TENSOR, EXPR)                       PLAN := plan(QUERIES...)
                                                QUERY := query(ALIAS, EXPR)
\end{minted}
\vspace{-12pt}
\caption{\footnotesize Finch Logic Syntax. }\label{fig:finch_logic_syntax}
\vspace{-12pt}
\end{wrapfigure}
\begin{figure}[b]
\begin{minted}{python}
  sum(A, dims=2) = aggregate(+, relabel(A, i_1, ..., i_d), i_2)
  matmul(A, dims=2) = aggregate(+, mapjoin(*, relabel(A, i, j), relabel(B, j, k)))
  argmin(A) = aggregate(minby, (Inf, len(A)), mapjoin(tuple, relabel(A, i), i), i)
  norm2(A) = aggregate(scaled_plus, (0, 0), mapjoin(tuple, mapjoin(sign, A), mapjoin(abs, A)), fields(A)...)
  where(C, A, B) = mapjoin(ifelse, relabel(C, i), relabel(A, i), relabel(B, i))
  slice(A, i:j) = table(WindowedArray(A, i:j), k)
  cat(A, B) = mapjoin(coalesce, table(PermissiveArray(A), i), table(OffsetArray(B, length(A)), i))
\end{minted}
\vspace{-12pt}
\caption{\footnotesize Implementations of a few common functions in Finch Logic. Here, \texttt{minby(x, y)} compares \texttt{x[1]} and \texttt{y[1]} and returns the smaller \texttt{x} or \texttt{y}. \texttt{scaled\_plus} rescales whichever argument is smaller, then adds the values with the same scale. \texttt{ifelse(c, a, b)} returns \texttt{a} when \texttt{c} is true,  \texttt{b} otherwise. Finch can understand declared properties and rewrite rules for all of these functions and types, extending these algorithms to operate on structured arguments.} \label{fig:api_extensible}
\vspace{-18pt}
\end{figure}
\subsubsection{Finch Logic and the Physical Interpreter}
Finch Logic is used as a common representation to implement and fuse the operations in the Array API.
%
The syntax is designed to be minimal and expressive, taking inspiration from relational algebra with the goal of being able to express most of the operations in the Array API.

\paragraph{Finch Logic}
The syntax of Finch Logic is described in Figure~\ref{fig:finch_logic_syntax}.
%
Our \texttt{mapjoin} represents pointwise application, while \texttt{aggregate} represents reductions.
%
We name the result of each computation with \texttt{query}, and join several expressions in a \texttt{plan}.
%
\texttt{reorder} reorders the indices while \texttt{relabel} relabels them.
%
\texttt{reformat} hints at materialization into its argument.

### 2024-04-05 sections/eval/spmv.tex
fixes

            y .= 0
            for j = _, i = _
              y[i] += A[i, j] * x[j]
            end
        \end{minted}
        \vspace{24pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            y .= 0
            for j = _, i = _
              y[j] += A[i, j] * x[i]
            end
        \end{minted}
    \end{minipage}\hfill%
    \begin{minipage}[t]{0.22\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            y .= 0
            for j = _
              let x_j = x[j]
                y_j .= 0
                for i = _
                  let A_ij = A[i, j]
                    y[i] += x_j * A_ij
                    y_j[] += A_ij * x[i]
                  end
                end
                #D is the diagonal
                y[j] += y_j[] + D[j] * x_j
              end
            end
        \end{minted}
    \end{minipage}
    \vspace{-12pt}
    \caption{Finch row-major, column-major and symmetric SpMV Programs}
    \label{spmv_programs}
    \vspace{-36pt}
  \end{wrapfigure}
  \subsection{Sparse Matrix-Vector Multiply (SpMV)}
  Sparse matrix-vector multiply (SpMV) has a wide range of applications and has been thoroughly studied~\cite{liu_csr5_2015,
  zhou_enabling_2020}. 
  %
  Because SpMV is bandwidth bound, many formats have
  been proposed to reduce the footprint~\cite{langr_evaluation_2016}. 
  %
  The wide range of applications unsurprisingly
  results in a wide range of array structures, making it an effective kernel to
  demonstrate the utility of our programming model. 
  %
  We varied both the data formats and the SpMV algorithm. 
  %
  Our formats
  are shown in Table~\ref{spmv_tensor_formats}, and our programs are listed in
  Figure~\ref{spmv_programs}.
  %
  In varying the program, we consider both row and column-major SpMV programs, as
  well as a symmetric SpMV. 
  %
  Finch enables us to exploit symmetry
  effectively.
  %
  Our program restricts our attention to the canonical triangle (using
  masks), reuses the reads to the canonical triangle of the symmetric matrix
  (using a $\finchdefine$ statement), and writes the results to both relevant
  locations (using multiple outputs).
  %
  All of our programs apply to all of our level formats, enabling exploitation of multiple structural patterns concurrently (e.g. sparsity \textit{and} symmetry).

  \begin{wrapfigure}{r}{0.5\linewidth}
    \scriptsize
    \centering
    \vspace{-16pt}
    \begin{tabular}{|r|l|l|}
        \hline
        \textbf{Format} & \textbf{Style of Matrix}\\
        \hline
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseList(Element(0.0)))} & real, sparse \\
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseList(Pattern()))} & boolean, sparse \\
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseVBL(Element(0.0)))} & real, blocked \\
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseBand(Element(0.0)))} & real, banded \\
        %\mintinline{julia}{Dense(SparsePoint(Element(0.0)))} & real-valued matrices with one value per row \\
        %\mintinline{julia}{Dense(SparseRLE(Pattern(0.0)))} & matrix with runs of true or false \\
        \hline 
    \end{tabular}
      \vspace{-8pt}
    \caption{SpMV Tensor Formats}
    \label{spmv_tensor_formats}
    \vspace{-8pt}
  \end{wrapfigure}

@@

  We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
  %
  Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had slowdowns of 0.71, 0.53, and 0.92 relative to TACO.
  %
  Furthermore, the synthetic banded matrices we constructed performed the best with the SparseBand matrix, in particular with the large\_band and the medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had slowdowns of 0.84 and 0.51 relative to TACO.
  %
  There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean. 
  %
  For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
  %
  Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
  %
  Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
  %
  Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.

@@

  \begin{figure}
      \includegraphics[width=\linewidth]{Spmv_annotated.drawio.png}
      \vspace{-18pt}
      \caption{Performance of SpMV by Finch format.}
      \label{fig:spmv_grouped}
      \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested. Programs are from Figure~\ref{spmv_programs} and formats are from Figure~\ref{spmv_tensor_formats}. ``finch\_baseline'' is the faster of row major or column major ``SparseList''.
  \end{figure}

>>>>

            y .= 0
            for j = _, i = _
              y[i] += A[i, j] * x[j]
            end
        \end{minted}
        \vspace{24pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            y .= 0
            for j = _, i = _
              y[j] += A[i, j] * x[i]
            end
        \end{minted}
    \end{minipage}\hfill%
    \begin{minipage}[t]{0.22\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            y .= 0
            for j = _
              let x_j = x[j]
                y_j .= 0
                for i = _
                  let A_ij = A[i, j]
                    y[i] += x_j * A_ij
                    y_j[] += A_ij * x[i]
                  end
                end
                #D is the diagonal
                y[j] += y_j[] + D[j] * x_j
              end
            end
        \end{minted}
    \end{minipage}
    \vspace{-12pt}
    \caption{Finch row-major, column-major and symmetric SpMV Programs}
    \label{spmv_programs}
    \vspace{-12pt}
  \end{wrapfigure}
  \subsection{Sparse Matrix-Vector Multiply (SpMV)}
  Sparse matrix-vector multiply (SpMV) has a wide range of applications and has been thoroughly studied~\cite{liu_csr5_2015,
  zhou_enabling_2020}. 
  %
  Because SpMV is bandwidth bound, many formats have
  been proposed to reduce the footprint~\cite{langr_evaluation_2016}. 
  %
  The wide range of applications unsurprisingly
  results in a wide range of array structures, making it an effective kernel to
  demonstrate the utility of our programming model. 
  %
  We varied both the data formats and the SpMV algorithm. 
  %
  Our formats
  are shown in Table~\ref{spmv_tensor_formats}, and our programs are listed in
  Figure~\ref{spmv_programs}.
  %
  In varying the program, we consider both row and column-major SpMV programs, as
  well as a symmetric SpMV. 
  %
  Finch enables us to exploit symmetry
  effectively.
  %
  Our program restricts our attention to the canonical triangle (using
  masks), reuses the reads to the canonical triangle of the symmetric matrix
  (using a $\finchdefine$ statement), and writes the results to both relevant
  locations (using multiple outputs).
  %
  All of our programs apply to all of our level formats, enabling exploitation of multiple structural patterns concurrently (e.g. sparsity \textit{and} symmetry).

  \begin{wrapfigure}{r}{0.5\linewidth}
    \scriptsize
    \centering
    \vspace{-14pt}
    \begin{tabular}{|r|l|l|}
        \hline
        \textbf{Format} & \textbf{Style of Matrix}\\
        \hline
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseList(Element(0.0)))} & real, sparse \\
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseList(Pattern()))} & boolean, sparse \\
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseVBL(Element(0.0)))} & real, blocked \\
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseBand(Element(0.0)))} & real, banded \\
        %\mintinline{julia}{Dense(SparsePoint(Element(0.0)))} & real-valued matrices with one value per row \\
        %\mintinline{julia}{Dense(SparseRLE(Pattern(0.0)))} & matrix with runs of true or false \\
        \hline 
    \end{tabular}
      \vspace{-8pt}
    \caption{SpMV Tensor Formats}
    \label{spmv_tensor_formats}
    \vspace{-8pt}
  \end{wrapfigure}

@@

  \begin{figure}[!h]
    \includegraphics[width=\linewidth]{Spmv_annotated.drawio.png}
    \caption{\footnotesize Performance of SpMV by Finch format. The performance displayed for Finch on each dataset is the fastest among the formats we tested. Programs are from Figure~\ref{spmv_programs} and formats are from Figure~\ref{spmv_tensor_formats}. ``finch\_baseline'' is the faster of row major or column major ``SparseList''.}
    \label{fig:spmv_grouped}
  \end{figure}

  We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
  %
  Matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had minor slowdowns relative to TACO.
  %
  Furthermore, the synthetic banded matrices we constructed performed the best with the SparseBand matrix, in particular with the large\_band and the medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had minor slowdowns relative to TACO.
  %
  The Pattern format performed better than the Element format for representing the leaf values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean. 
  %
  For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
  %
  Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
  %
  Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
  %
  Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.

@@
